Explainability and Adversarial Robustness for RNNs

### 2019-12-20 paper/paper.tex
Fix leaving out features section

Network communication is usually aggregated into \textit{flows}, which are commonly defined as a sequence of packets sharing certain properties. %, between two applications potentially being located at different physical locations.
When analyzing flows, not only the aforementioned features are available but also features related to the timing of the individual packets. % as well as their size. %AH: don't see why pkt size is not available without flows
%For example, a specific attack might require the attacker to send three small packets and the victim to reply with two large packets followed by a small one to the attacker.
Various approaches have been proposed to extract features from flows and then perform anomaly detection with the extracted flows \cite{meghdouri_analysis_2018}.
While these approaches frequently work well, it is problematic that the whole flow has to be received first and only afterwards anomaly detection is applied, revealing attack flows. Thus, we design a network \gls{ids} that operates on a per-packet basis and decides if a packet is anomalous based on features that are available even for traffic that is encrypted above the transport layer, like for example TLS or QUIC.
At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding tedious feature engineering procedures, which derive statistical measures from the sequence of packet features, but providing any available information to the classifier.
We show that our system has similar performance to other flow-based anomaly detection systems but can detect anomalies \textit{before} the flow terminates.

@@

\begin{table}[b]
\caption{Flow occurrence frequency of attack types.}
\label{tab:occurrence}
\centering
\subfloat[CIC-IDS-2017\label{fig:cicids17proportions}
]{
\begin{tabular}{l r}
\toprule
Attack type & \hspace*{-4mm}Proportion \\ \midrule
DoS Hulk	&	10.10\%	\\
PortScan, Firewall	&	6.90\%	\\
DDoS LOIT	&	4.08\%	\\
Infiltration	&	3.30\%	\\
DoS GoldenEye	&	0.32\%	\\
DoS SlowHTTPTest	&	0.18\%	\\
DoS Slowloris	&	0.17\%	\\
Brute-force SSH	&	0.11\%	\\
Botnet ARES	&	0.03\%	\\
XSS attack	&	0.03\%	\\
PortScan, no Fw.	&	0.02\%	\\
Brute-force FTP	&	0.01\%	\\
SQL injection	&	$<$0.01\%	\\
Heartbleed	&	$<$0.01\%	\\
\bottomrule
\end{tabular}
}{}
\subfloat[UNSW-NB15\label{fig:unswnb15proportions}
]{
\begin{tabular}{l r}
\toprule
Attack type & \hspace*{-4mm}Proportion \\ \midrule
Exploits	&	1.42\%	\\
Fuzzers	&	1.01\%	\\
Reconnaissance	&	0.58\%	\\
Generic	&	0.21\%	\\
DoS	&	0.19\%	\\
Shellcode	&	0.08\%	\\
Analysis	&	0.03\%	\\
Backdoors	&	0.02\%	\\
Worms	&	0.01\%	\\
\bottomrule
\end{tabular}
}{}
\end{table}

We implemented a three-layer LSTM-based classifier with 512 neurons at each layer. As the input features we use
source port, destination port, protocol identifier, packet length, \gls{iat} to the previous packet in the flow, packet direction (i.e. forward or reverse path) and all TCP flags (0 if the flow is not TCP).
% \begin{itemize*}
% \item source port
% \item destination port
% \item protocol identifier
% \item packet sizes
% \item \glspl{iat} of packets
% \item packet directions (if packet goes there or back)
% \item all TCP flags (0 if the flow is not TCP)
% \end{itemize*}.
We omitted \gls{ttl} values, as they are likely to lead to unwanted prediction behaviour \cite{bachl_walling_2019}.  Among the used features, source port, destination port and protocol identifier are constant over the whole flow while the others vary.
During flow extraction we used the usual 5-tuple flow key, which distinguishes flows based on the protocol they use and their source and destination port and IP address.
We use Z-score normalization and a train/test split of 2:1.

@@

As \autoref{fig:adv_per_family} shows, for most attack types, most samples can be modified by an adversary to be classified as benign by using \gls{cw}. High recall for a particular attack type does not imply that adversarial samples are hard to find  %are  likely to fail
for these attacks. Thus we argue that beside the classical metrics such as accuracy, false positives etc., a metric is needed, which quantifies how easily an \gls{ids} classifier can be fooled.

@@

\begin{figure}[h]
\includegraphics[width=\columnwidth]{../plots/ars_original.pdf}
\caption{Recall for unmodified flows and \gls{ars} for each attack type in CIC-IDS-2017. For ``PortScan, Firewall'' recall never falls under 66\%: The adversary does not succeed and the \gls{ars} is $\infty$.}
\label{fig:recall_ars}
\end{figure}

@@

By using a classical \gls{pdp} we would likely lose information due to the averaging over very different types of traffic. However, for network traffic in particular, investigating each sample individually is not possible. Hence, the conditional \gls{pdp} provides the ideal compromise for our application.

@@

Both approaches lead to complete resistance to adversarial samples (as we left out all manipulable features), except for botnets, which can still operate when only leaving out manipulable features in one direction. The results show that -- surprisingly -- there is only a small difference in classification performance when only computing the accuracy for the last packet of each flow (flow accuracy): Flow accuracy stays over 99.93\%. However, packet accuracy is only 98\% when leaving out the features in one direction and 96.7\% when leaving them out in both directions. Thus, apparently the \gls{iat} and packet size are especially important for determining whether a flow is malicious in the first packets of a flow.

>>>>

Network communication is usually aggregated into \textit{flows}, which are commonly defined as a sequence of packets sharing certain properties. %, between two applications potentially being located at different physical locations.
When analyzing flows, not only the aforementioned features are available but also features related to the timing of the individual packets. % as well as their size. %AH: don't see why pkt size is not available without flows
%For example, a specific attack might require the attacker to send three small packets and the victim to reply with two large packets followed by a small one to the attacker.
Various approaches have been proposed to extract features from flows and then perform anomaly detection with the extracted flows \cite{meghdouri_analysis_2018}.
While these approaches frequently work well, it is problematic that the whole flow has to be received first and only afterwards anomaly detection is applied, revealing attack flows. Thus, we design a network \gls{ids} that operates on a per-packet basis and decides if a packet is anomalous based on features that are available even for traffic that is encrypted above the transport layer, like for example TLS or QUIC.
At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding tedious feature engineering procedures, which derive statistical measures from the sequence of packet features, but providing any available information to the classifier.
We show that our system has similar performance to other flow-based anomaly detection systems but can detect anomalies \textit{before} the flow is over.

@@

\begin{table}[b]
\caption{Occurrence frequency of attack types.}
\label{tab:occurrence}
\centering
\subfloat[CIC-IDS-2017\label{fig:cicids17proportions}
]{
\begin{tabular}{l r}
\toprule
Attack type & \hspace*{-4mm}Proportion \\ \midrule
DoS Hulk	&	10.10\%	\\
PortScan, Firewall	&	6.90\%	\\
DDoS LOIT	&	4.08\%	\\
Infiltration	&	3.30\%	\\
DoS GoldenEye	&	0.32\%	\\
DoS SlowHTTPTest	&	0.18\%	\\
DoS Slowloris	&	0.17\%	\\
Brute-force SSH	&	0.11\%	\\
Botnet ARES	&	0.03\%	\\
XSS attack	&	0.03\%	\\
PortScan, no Fw.	&	0.02\%	\\
Brute-force FTP	&	0.01\%	\\
SQL injection	&	$<$0.01\%	\\
Heartbleed	&	$<$0.01\%	\\
\bottomrule
\end{tabular}
}{}
\subfloat[UNSW-NB15\label{fig:unswnb15proportions}
]{
\begin{tabular}{l r}
\toprule
Attack type & \hspace*{-4mm}Proportion \\ \midrule
Exploits	&	1.42\%	\\
Fuzzers	&	1.01\%	\\
Reconnaissance	&	0.58\%	\\
Generic	&	0.21\%	\\
DoS	&	0.19\%	\\
Shellcode	&	0.08\%	\\
Analysis	&	0.03\%	\\
Backdoors	&	0.02\%	\\
Worms	&	0.01\%	\\
\bottomrule
\end{tabular}
}{}
\end{table}

We implemented a three-layer LSTM-based classifier with 512 neurons at each layer. As the input features we use
source port, destination port, protocol identifier, packet length, \gls{iat} to the previous packet in the flow, packet direction (if packet goes there or back) and all TCP flags (0 if the flow is not TCP).
% \begin{itemize*}
% \item source port
% \item destination port
% \item protocol identifier
% \item packet sizes
% \item \glspl{iat} of packets
% \item packet directions (if packet goes there or back)
% \item all TCP flags (0 if the flow is not TCP)
% \end{itemize*}.
We omitted \gls{ttl} values, as they are likely to lead to unwanted prediction behaviour \cite{bachl_walling_2019}.  Among the used features, source port, destination port and protocol identifier are constant over the whole flow while the others vary.
During flow extraction we used the usual 5-tuple flow key, which distinguishes flows based on the protocol they use and their source and destination port and IP address.
We use Z-score normalization and a train/test split of 2:1.

@@

As \autoref{fig:adv_per_family} shows, for most attack types, most samples can be modified by an adversary to be classified as benign by using \gls{cw}. High accuracy for a particular attack category does not imply that adversarial attacks are  likely to fail for these attacks. Thus we argue that beside the classical metrics such as accuracy, false positives etc., a metric is needed, which quantifies how easily an \gls{ids} classifier can be fooled.

@@

\begin{figure}[h]
\includegraphics[width=\columnwidth]{../plots/ars_original.pdf}
\caption{Recall for each attack type and \gls{ars} for CIC-IDS-2017. For ``PortScan, Firewall'' recall never falls under 66\%: The adversary does not succeed and the \gls{ars} is $\infty$.}
\label{fig:recall_ars}
\end{figure}

@@

By using a classical \gls{pdp} we would be likely to lose information due to the averaging over very different types of traffic. However, for network traffic in particular, investigating each sample individually is not possible. Hence, the conditional \gls{pdp} provides the ideal compromise for our application.

@@

Both approaches lead to complete resistance to adversarial samples, except for botnets, which can still operate when only leaving out manipulable features in one direction. The results show that -- surprisingly -- there is only a small difference in classification performance when looking at flow accuracy: It is slightly lower at 99.3\% compared to 99.7\% originally for CIC-IDS-2017. However, packet accuracy is only 98\% when leaving out the features in one direction and 96.7\% when leaving them out in both directions. Thus, apparently the \gls{iat} and packet size are especially important for determining whether a flow is malicious in the first packets of a flow.
